A Reduced Order Model for Permeability Fields Using Singular Value Decomposition (SVD)

### 2018-04-13 Papers/AppliedMaths2018/Stochastic_Upscaling_UK_AMM_2018.tex
updated upscaling (Section 3)

\red{Motivation for the work? What is the main aim of the work?} In this paper, a SVD-based method was developed to reduce the order of porous media permeability field (\ie upscale). This method was compared to traditional industry-standard upscaling techniques (\eg arithmetic and harmonic averaging) and to a stochastic-based (probability density function, PDF) method of upscaling, to highlight its benefits \red{(performance??)}. It was shown that the SVD upscaling technique retained the heterogenous nature of the BaseCase (\ie high-resolution configuration). Additionally, the SVD upscaling technique did not require {\it a priori} understanding of similar permeability blocks within the existing domain.

@@

\begin{keyword} %% keywords here, in the form: keyword \sep keyword
Singular Value Decomposition (SVD)\sep Upscaling\sep Model Order Reduction (MOR) \sep  Principal Component Spaces \sep Permeability field
\end{keyword}

@@

\citet{Schilders2008} provided a historical perspective on the development of Model Order Reduction, from the work of Fourier in 1807 which approximated a function with a few trigonometric terms, to the work of \citet{Odabasioglu_1998} in 1998 which introduced Passive Reduced-order Interconnect Macromodeling Algorithm (PRIMA). Truncated Balanced Realization  is another MOR method developed by \citeauthor{Moore_1981}~\cite{Moore_1981}, which introduced the principal component analysis \cite{Hotelling_1933} and \citeauthor{Golub1970}'s \cite{Golub1970} algorithm for solving singular value decomposition (SVD). Hankel-norm reduction \cite{Glover_1984} and Proper Orthogonal Decomposition \cite{Sirovich_1987} are also MOR methods developed published in 1984 and 1987 respectively\red{(I DONT THINK THIS IS ACCURATE ... POD OR KARHUNEN-LOEVE DECOMPOSITION/TRANSFORM ORIGINAL PAPERS ARE FROM 1933-48!)}. The asymptotic waveform evaluation (AWE) \cite{Pillage_1990} was the first MOR method based on Krylov subspace technique, this method was followed by Pade-via-Lanczos method \cite{Feldmann_1995} and the PRIMA. In this work, upscaling is performed by SVD in principal component spaces.

\bigskip
This work focuses on reducing the order of permeability distribution in porous media flow simulations. Section~\ref{section:mathematical_model} introduces the mathematical model used to describe the relevant physics for multi-fluid flow through porous media and highlights the relevance of the permeability field. The mathematical model is discretised by a novel high-order accurate control volume finite element method (CVFEM) \cite{gomes_2017} which couples both the fine resolution and upscaled representation of the permeability field. Pre-processing statistical properties and post-simulation multiphase flow behaviour are investigated to assess the various upscaling techniques used.

@@

Section~\ref{section:intro} briefly introduced the rational behind upscaling. Upscaling methods replace discrete geological and fluid properties of detailed high resolution domains with coarse descriptions (low resolution) of these properties \cite{Vereecken_2007}. This is a subject which has interested the academic community for over $7$ decades.  \citet{Cardwell_1945} investigated arithmetic and weighted averages as upscaling techniques for calculating a single equivalent permeability for an oil reservoir with varying permeabilities. \citet{King1996} investigated the use of renormalization for upscaling. And \citet{Yeo2001} discussed the accuracy of renormalization method for upscaling 2D hydraulic conductivity fields. \citet{Renard_1997} provided a comprehensive review of the various techniques for upscaling. \citet{Christie_2001SPE10Model} provided $2$ sets of problems to compare the upscaling and upgridding techniques of different simulators. \red{(And what are the main conclusions/contributions of these papers?)}

\JGnote{You should consider writing a more detailed section on upscaling as this is part of the core of your work here ...}

\BLnote{I will do this before submitting, please let me know if you will be able to review again on Thursday or Friday}

\section{Singular Value Decomposition, SVD, and its application for reducing the order of a permeability field}\label{section:svd}
\subsection{A Brief Introduction to SVD}\label{subsection:svd_brief}
Given a matrix $\mathbf{A}\in\mathcal{R}^{m\times n}$, the singular value decomposition (SVD) is a method for factorizing $\mathbf{A}$, such that,
\begin{equation}
  \mathbf{A} = \mathbf{U}\mathbf{\Sigma}\mathbf{V}^{\intercal}.
\end{equation}
where $\mathbf{U}=\left[\underline{u}_{1},\underline{u}_{2},\cdots,\underline{u}_{m}\right]\in\mathcal{R}^{m\times n}$ and $\mathbf{V}=\left[\underline{v}_{1},\underline{v}_{2},\cdots,\underline{v}_{n}\right]\in\mathcal{R}^{m\times n}$ are orthogonal but not necessarily the same. $\Sigma$ is a diagonal matrix, \ie $\Sigma=\text{diag}\left(\sigma_{1},\sigma_{2},\cdots,\sigma_{r}\right)\in\mathcal{R}^{m\times n}$ and $r=\min\left\{m,n\right\}$ is the rank of $\mathbf{A}$, with $\sigma_{1}\ge\sigma_{2}\ge\cdots\sigma_{r}\ge 0$. The diagonal entries of $\Sigma$ $\left(\ie \sigma_{i}, i=1,r\right)$ are called singular values of $\mathbf{A}$. If $\mathbf{A}$ is a square matrix then $\mathbf{U}$, $\Sigma$ and $\mathbf{V}$ are also square matrices. Also, as $\mathbf{U}$ and $\mathbf{V}$ are orthogonals and unitary, 
\begin{equation}
  \mathbf{U}^{\intercal}\mathbf{U} = \mathbf{V}^{\intercal}\mathbf{V} = \mathbf{I},
\end{equation}
where $\mathbf{I}$ is an identity matrix. However, if $\mathbf{A}$ is an $m\times n$ matrix, $\mathbf{U}$ and $\mathbf{V}$ still remain as square matrices, albeit with different matricial sizes. $\mathbf{U}$ is an $m\times m$ while $\mathbf{V}$ is an $n\times n$. Another relevant property of $\mathbf{U}$ and $\mathbf{V}$ since they are square (irrespective of whether $\mathbf{A}$ is square or rectangular) is that
\begin{equation}
  \mathbf{U}^{-1} = \mathbf{U}^{\intercal} \;\;\;\text{ and }\;\;\; \mathbf{V}^{-1} = \mathbf{V}^{\intercal},
\end{equation}
and if $\mathbf{A}$ is an $m \times n$ matrix thus $\Sigma$ is a diagonal $m \times n$ matrix. For a rectangular matrix, the singular values fill the first $r$ places on the main diagonal of $\Sigma$. The SVD benefits from the orthogonality of the columns in $\mathbf{U}$ and $\mathbf{V}$. The columns of $\mathbf{U}$ are known as the left singular vectors while the columns in $\mathbf{V}$ are known as the right singular vectors. A simple consideration of the benefit of the orthogonal factors, $\mathbf{U}$ and $\mathbf{V}$, is the ease with which they allow the inverse of the square matrix $\mathbf{A}$ to be calculated. This in turn makes it easy to solve the linear Equ.~\ref{e:Ax=b} for $x$. Equ.~\ref{e:SVD_factorization} show the SVD factorization already discussed, and with benefit of the orthogonal property also already discussed, the solution for $x$ is simply presented as in Equ.~\ref{e:x_solution}.
\begin{equation}\label{e:Ax=b}
 \mathbf{A}x = b
\end{equation}
\begin{equation}\label{e:SVD_factorization}
 \mathbf{U} \Sigma \mathbf{V}^{\intercal} x = b
\end{equation}
\begin{equation}\label{e:x_solution}
 x = \mathbf{V} \Sigma^{-1} \mathbf{U}^{\intercal} b
\end{equation}
\red{(I guess we have talked about this on Friday ... make it clear that matrix inversion is an (useful?) example of SVD ... also explain step-by-step this matricial equations above ... remember ... every single equation must be explained ...)}

\BLnote{The equations are now explained in the last 3 sentences of this section.}

\subsection{Principal Component Spaces and Model Order Reduction for the permeability field}\label{subsection:pcspaces}
\JGnote{As we talked on Friday: (a)make sure you explain what permeability is before starting to operating on it; (b) $P0DG$ could be easily replaced by saying that the quantity of interest is allocated in the centre of the element/cell -- this make it more general; (c) Just reading it through, we can summarise the {\it procedure} as (i) SVD of matrix of permeability scalar quantities (\ie upscaling) and (ii) linear interpolation of the upscaled/homogeneised quantities upon a new mesh grid resolution (\ie upgriding). Whereas the first is the object of the interest for this study, the second gives an overall motivation for the work ... thus, make sure this is clear within the text, \ie upscaling $\rightarrow$ upgriding.}

\BLnote{(a)Permeability is now introduced in Section~\ref{section:mathematical_model}; (b) $P0DG$ now replaced with centre of element/cell (c) I thought we agreed that upscaling cannot be separated from upgridding. Actually, you convinced me that upscaling cannot be seperated from upgridding. Upscaling is not completed in the SVDCase until the interpolation is completed. Strictly speaking, interpolation only defines the level of upgridding. I will mention this in the results section (Section~\ref{section:results_discussion})}

@@

\JGnote{Maybe you should consider describing the procedure above as pseudocode within a table.}

\BLnote{I have an algorithm in Section~\ref{subsubsection:svdcase_preprocess_algorithm} which itemizes the procedure in a sort of psuedocode}

@@

\subsection{Model Description}\label{subsection:model}
The BaseCase  ({\it i.e.} fine resolution) model for the permeabiliy field builds on the $2 * 2$ block, $4$ quadrant model used initially by \citet{Cardwell_1945} and later by \citet{Yeo2001} and \citet{dawe_2008}. These authors designed the permeability field for the quadrants in a checkerboard style which means that diagonally opposite blocks retained the same permeability values with one set of diagonally opposite blocks having a higher permeability value compared to the other set. Hence, one set of diagonally opposite blocks are referred to as high permeability blocks while the other set is referred to as low permeability blocks. The BaseCase permeability model used in this manuscript retains the checkerboard desgined used by the recently named authors. Albeit, the permeability data for each block in the set of diagonally opposite blocks have similar value ({\it i.e.} not exactly the same as in the papers referred to), although one set can still be referred to as a set of high permeability blocks while the other set is referred to as a set of low permeability blocks. Furthermore, within each block, the permeability field varies within a pre-defined range and consists of $1600$ permeability values. More details on the design of the permeability field for the BaseCase is available in section~\ref{subsubsection:basecase_preprocess_algorithm}. And Fig.~\ref{fig:PermFields}a and~\ref{fig:PermFields}b give a pictorial representation of the permeability field for the base case with the mesh discretization visible and invisible respectively. Using the labelling scheme used by \citet{dawe_2008}, the quadrant model block $K1,~K2,~K3$ and $K4$ are top left, bottom left, top right and bottom right respectively.

>>>>

Representation of heterogeneous properties (\ie permeability in this case) within a large system, is a challenge that cuts across many industries that deal with flow in porous media. This is because it is impossible to obtain the value of permeability at every point in realistic domain. And, even if this was possible, it would be extremely challenging to get the computing resources to make use of all this data. This challenge gave birth to upscaling, a model order reduction method.

In this paper, a SVD-based method was developed to reduce the order of porous media permeability field (\ie upscale). This method was compared to traditional industry-standard upscaling techniques (\eg arithmetic and harmonic averaging) and to a stochastic-based (probability density function, PDF) method of upscaling, to highlight its benefits performance. It was shown that the SVD upscaling technique retained the heterogenous nature of the BaseCase (\ie high-resolution configuration). Additionally, the SVD upscaling technique did not require {\it a priori} understanding of similar permeability blocks within the existing domain.

@@

\begin{keyword} %% keywords here, in the form: keyword \sep keyword
Singular Value Decomposition (SVD)\sep Upscaling\sep Model Order Reduction (MOR) \sep  Principal Component Spaces \sep Permeability field \sep Heterogeneity
\end{keyword}

@@

\citet{Schilders2008} provided a historical perspective on the development of Model Order Reduction, from the work of Fourier in 1807 which approximated a function with a few trigonometric terms, to the work of \citet{Odabasioglu_1998} in 1998 which introduced Passive Reduced-order Interconnect Macromodeling Algorithm (PRIMA). Truncated Balanced Realization  is another MOR method developed by \citeauthor{Moore_1981}~\cite{Moore_1981}, which introduced the principal component analysis \cite{Hotelling_1933} and \citeauthor{Golub1970}'s \cite{Golub1970} algorithm for solving singular value decomposition (SVD). Hankel-norm reduction \cite{Glover_1984} and Proper Orthogonal Decomposition \cite{Sirovich_1987} are also MOR methods discussed in the referenced articles published in 1984 and 1987 respectively. The asymptotic waveform evaluation (AWE) \cite{Pillage_1990} was the first MOR method based on Krylov subspace technique, this method was followed by Pade-via-Lanczos method \cite{Feldmann_1995} and the PRIMA. In this work, upscaling is performed by SVD in principal component spaces.

\bigskip
This work focuses on reducing the order of permeability distribution in porous media flow simulations. Section~\ref{section:mathematical_model} introduces the mathematical model used to describe the relevant physics for multi-fluid flow through porous media and highlights the relevance of the permeability field. The mathematical model is discretised by a novel high-order accurate control volume finite element method (CVFEM) \cite{Gomes_2017} which couples both the fine resolution and upscaled representation of the permeability field. Pre-processing statistical properties and post-simulation multiphase flow behaviour are investigated to assess the various upscaling techniques used.

@@

Section~\ref{section:intro} briefly introduced the rational behind upscaling. Upscaling methods replace discrete geological and fluid properties of detailed high resolution domains with coarse descriptions (low resolution) of these properties \cite{Vereecken_2007}. This subject of upscaling heterogenous properties has interested industries (\ie oil and gas, CCS and waste management) as well as the academic community for over $7$ decades.

Those interested in large-scale reservoirs or ground water generally take a two-stage geostatistical approach to be subject. In the first stage, information from seismic data, well data and analogous outcrops is used to model the large scale heterogeneities associated with facies. And in the second stage, statistical models such as continuous multi-variate Gaussian field is used to model the rock properties of the facies \cite{Ewing_1997}. In addition to seismic and well data used in the previous stage, core data can be used as a sources of information for the statistical properties used to determine the random field in this stage.

A lot of work has been carried out on the upscaling of permeability fields \cite{Christie_2001SPE10Model, Ewing_1997, Hasting_2001, King1996, Indelman_1993, Vereecken_2007, Wen_1996, Yeo2001}. \citet{Cardwell_1945} investigated arithmetic and weighted averages as upscaling techniques for calculating a single equivalent permeability for an oil reservoir with varying permeabilities. The general conclusion reached from this study was that the equivalent permeability lies between the harmonic and the arithmetic average.

\citet{Yeo2001} discussed the accuracy of renormalization method for upscaling 2D hydraulic conductivity fields for two cases. Both cases employed the $4$-quadrant model (see Section~\ref{subsection:model} for details) set-up. The first case had low permeablitity in all the blocks except the bottom right block which had high permeability, while the second case employed the checkboard style described in Section~\ref{subsection:model}. Different conductivity ratios were also investigated for the two cases. The conclusion drawn was that renormalization worked fairly well for the first case, but very poorly for the second case.

\citet{Renard_1997} provided a comprehensive review of the various techniques for upscaling. He found out that the techniques classed broadly under deterministic, stochastic and heuristic complemented each other rather than antagonising each other. \citet{Renard_1997} then recommended that it is better to select a technique which provides block permeabilities instead of one which provides uniform effective permeability.

\citet{Christie_2001SPE10Model} provided $2$ sets of problems to compare the upscaling and upgridding techniques of different simulators. The paper does not describe in any detail the upscaling techniques used by the different simulators, but it provides a complete dataset for running simulations.

\section{Singular Value Decomposition, SVD, and its application for reducing the order of a permeability field}\label{section:svd}
\subsection{A Brief Introduction to SVD}\label{subsection:svd_brief}
Given a matrix $\mathbf{A}\in\mathcal{R}^{m\times n}$, the singular value decomposition (SVD) is a method for factorizing $\mathbf{A}$, such that,
\begin{equation}
  \mathbf{A} = \mathbf{U}\mathbf{\Sigma}\mathbf{V}^{\intercal}.
\end{equation}
where $\mathbf{U}=\left[\underline{u}_{1},\underline{u}_{2},\cdots,\underline{u}_{m}\right]\in\mathcal{R}^{m\times n}$ and $\mathbf{V}=\left[\underline{v}_{1},\underline{v}_{2},\cdots,\underline{v}_{n}\right]\in\mathcal{R}^{m\times n}$ are orthogonal but not necessarily the same. $\Sigma$ is a diagonal matrix, \ie $\Sigma=\text{diag}\left(\sigma_{1},\sigma_{2},\cdots,\sigma_{r}\right)\in\mathcal{R}^{m\times n}$ and $r=\min\left\{m,n\right\}$ is the rank of $\mathbf{A}$, with $\sigma_{1}\ge\sigma_{2}\ge\cdots\sigma_{r}\ge 0$. The diagonal entries of $\Sigma$ $\left(\ie \sigma_{i}, i=1,r\right)$ are called singular values of $\mathbf{A}$. If $\mathbf{A}$ is a square matrix then $\mathbf{U}$, $\Sigma$ and $\mathbf{V}$ are also square matrices. Also, as $\mathbf{U}$ and $\mathbf{V}$ are orthogonals and unitary, 
\begin{equation}
  \mathbf{U}^{\intercal}\mathbf{U} = \mathbf{V}^{\intercal}\mathbf{V} = \mathbf{I},
\end{equation}
where $\mathbf{I}$ is an identity matrix. However, if $\mathbf{A}$ is an $m\times n$ matrix, $\mathbf{U}$ and $\mathbf{V}$ still remain as square matrices, albeit with different matricial sizes. $\mathbf{U}$ is an $m\times m$ while $\mathbf{V}$ is an $n\times n$. Another relevant property of $\mathbf{U}$ and $\mathbf{V}$ since they are square (irrespective of whether $\mathbf{A}$ is square or rectangular) is that
\begin{equation}
  \mathbf{U}^{-1} = \mathbf{U}^{\intercal} \;\;\;\text{ and }\;\;\; \mathbf{V}^{-1} = \mathbf{V}^{\intercal},
\end{equation}
and if $\mathbf{A}$ is an $m \times n$ matrix thus $\Sigma$ is a diagonal $m \times n$ matrix. For a rectangular matrix, the singular values fill the first $r$ places on the main diagonal of $\Sigma$. The SVD benefits from the orthogonality of the columns in $\mathbf{U}$ and $\mathbf{V}$. The columns of $\mathbf{U}$ are known as the left singular vectors while the columns in $\mathbf{V}$ are known as the right singular vectors. A simple consideration of the benefit of the orthogonal factors, $\mathbf{U}$ and $\mathbf{V}$, is the ease with which they allow the inverse of the square matrix $\mathbf{A}$ to be calculated. This in turn makes it easy to solve the linear Equ.~\ref{e:Ax=b} for $x$. Equ.~\ref{e:SVD_factorization} show the SVD factorization already discussed, and with benefit of the orthogonal property also already discussed, the solution for $x$ is simply presented as in Equ.~\ref{e:x_solution}.
\begin{equation}\label{e:Ax=b}
 \mathbf{A}x = b
\end{equation}
\begin{equation}\label{e:SVD_factorization}
 \mathbf{U} \Sigma \mathbf{V}^{\intercal} x = b
\end{equation}
\begin{equation}\label{e:x_solution}
 x = \mathbf{V} \Sigma^{-1} \mathbf{U}^{\intercal} b
\end{equation}

\subsection{Principal Component Spaces and Model Order Reduction for the permeability field}\label{subsection:pcspaces}

@@

@@

\subsection{Model Description}\label{subsection:model}
The BaseCase  ({\it i.e.} fine resolution) model for the permeabiliy field builds on the $2 \times 2$ block, $4$ quadrant model used initially by \citet{Cardwell_1945} and later by \citet{Yeo2001} and \citet{dawe_2008}. These authors designed the permeability field for the quadrants in a checkerboard style which means that diagonally opposite blocks retained the same permeability values with one set of diagonally opposite blocks having a higher permeability value compared to the other set. Hence, one set of diagonally opposite blocks are referred to as high permeability blocks while the other set is referred to as low permeability blocks. The BaseCase permeability model used in this manuscript retains the checkerboard desgined used by the recently named authors. Albeit, the permeability data for each block in the set of diagonally opposite blocks have similar value ({\it i.e.} not exactly the same as in the papers referred to), although one set can still be referred to as a set of high permeability blocks while the other set is referred to as a set of low permeability blocks. Furthermore, within each block, the permeability field varies within a pre-defined range and consists of $1600$ permeability values. More details on the design of the permeability field for the BaseCase is available in section~\ref{subsubsection:basecase_preprocess_algorithm}. And Fig.~\ref{fig:PermFields}a and~\ref{fig:PermFields}b give a pictorial representation of the permeability field for the base case with the mesh discretization visible and invisible respectively. Using the labelling scheme used by \citet{dawe_2008}, the quadrant model block $K1,~K2,~K3$ and $K4$ are top left, bottom left, top right and bottom right respectively.
